---
title: "Chapter 2. Time series models, trend and autocovariance"
output: html_notebook
---

### Trend
Define the mean function (or trend) for $n \in 1: N$ by
\[
\mu_{n}=\mathbb{E}\left[Y_{n}\right]=\int_{-\infty}^{\infty} y_{n} f_{Y_{n}}\left(y_{n}\right) d y_{n}
\]
We only make observations at the discrete collection of times $t_1:N$ and so we require $\mu(t_n) = \mu_n$.

If the mean is constant, that is, $\mu_n = \mu$, the model is called **mean stationary**. We can estimate mean use the mean estimator, $\hat{\mu} = \hat{\mu}\left(y_{1: N}\right)=\frac{1}{N} \sum_{k=1}^{N} y_{k}$. It is only a reasonable estimator of the mean function when a mean stationary model is appropriate.

Trend is a property of the *model*, and the estimated trend is a function of the *data*.

### The autocovariance function
The covariance between two random variables $Y_m$ and $Y_n$
\[
\gamma_{m, n}=\mathbb{E}\left[\left(Y_{m}-\mu_{m}\right)\left(Y_{n}-\mu_{n}\right)\right]
\]
If the covariance between two observations **depends only on their time difference**, the time series model is **covariance stationary**. For observations equally spaced in time, the autocovariance function is then a function of a lag $h$, $\gamma_{h}=\gamma_{n, n+h}$.

For a covariance stationary model, a common *estimator* for $\gamma_h$ is
\[
\hat{\gamma}_{h}\left(y_{1: N}\right)=\frac{1}{N} \sum_{n=1}^{N-h}\left(y_{n}-\hat{\mu}_{n}\right)\left(y_{n+h}-\hat{\mu}_{n+h}\right)
\]
where $\hat{\mu}_{n}=\hat{\mu}_{n}\left(y_{1: N}\right)$ is a mean function estimator.

### The autocorrelation function
Dividing the autocovariance by the variance is the autocorrelation function $\rho_h$ given by
\[
\rho_{h}=\frac{\gamma_{h}}{\gamma_{0}}
\]
The standard autocorrelation estimator is given by $\hat{\rho}_{h}\left(y_{1: N}\right)=\frac{\hat{\gamma}_{h}\left(y_{1: N}\right)}{\hat{\gamma}_{0}\left(y_{1: N}\right)}$ and results in an estimate, the sample autocorrelation $\hat{\rho}_{h}=\hat{\rho}_{h}\left(y_{1: N}\right)=\frac{\hat{\gamma}_{h}}{\hat{\gamma}_{0}}$.

Note that **correlation** and **covariance** are properties of models. Sample autocorrelation and sample autocovariance are related to data.

```{r}
global_temp <- read.table("data/Global_Temperature.txt",header=TRUE)
str(global_temp)
```
```{r}
plot(Annual~Year,data=global_temp,ty="l")
```

### Fitting a least squares model with a quadratic trend

```{r}
lm_fit <- lm(Annual~Year+I(Year^2),data=global_temp)
```
here `I()` is a function that tells R to construct `Year^2` as a variable, and inhibits interpretation in the R model formula notation.
```{r}
summary(lm_fit)
```
Check visually how well this model fits the data:
```{r}
yr <- 1880:2026
Z <- cbind(1,yr,yr^2)
beta <- coef(lm_fit)
prediction <- Z%*%beta
plot(Annual~Year,data=global_temp,ty="l",xlim=range(yr),
ylim=range(c(global_temp$Annual,prediction),na.rm=TRUE),
  lty="dashed")
lines(x=yr,y=prediction,col="red")
```

### White noise error model
The usual model behind OLS is the independent error model, known in time series analysis as the **white noise error model**.

A result for linear models is that $\hat{\beta}_{O L S}\left(y_{1: N}\right)$ is the minimum variance unbiased estimator for model $Y = Z\beta + \epsilon$, where $\epsilon = \epsilon_{1:N}$ is a vector of independent, identically distributed random variables with mean zero and constant variance,$\mathbb{E}\left[\epsilon_{n}\right]=0,  \operatorname{Var}\left[\epsilon_{n}\right]=\sigma^{2}$. 

`lm` in R, provides confidence intervals based on this model.

The variance/covariance matrix of $\hat{\beta}_{O L S}\left(Y_{1: N}\right)$ under this model is 
\[
\operatorname{Cov}\left[\hat{\beta}_{O L S}\left(Y_{1: N}\right)\right]=\sigma^{2}\left(Z^{\mathrm{T}} Z\right)^{-1}
\]
and is estimated using an estimator for $\sigma$ by 
\[
\hat{\sigma}_{O L S}\left(y_{1: N}\right)=\sqrt{\frac{1}{N-d}\left(y-Z \hat{\beta}_{O L S}\right)^{\mathrm{T}}\left(y-Z \hat{\beta}_{O L S}\right)}
\]
$d$ is the number of covariates, i.e., the number of columns of $Z$.
```{r}
acf(resid(lm_fit))
```

The *horizontal dashed lines* on the graph of the sample autocorrelation function (ACF) give **a measure of chance variation under the null hypothesis that the residuals are IID**.

At each lag h, the chance that the estimated ACF falls within this band is approximately $95\%$. Thus, under the null hypothesis, one expects a fraction of $1/20$ of the lags of the sample ACF to fall outside this band.

However, the sample ACF confirms what we can probably see from the plot of the fitted model: the variation around the fitted model is **clustered in time**, so the sample ACF of the residuals is not consistent with a model having independent error terms.

### Generalized least squares

If we know the covariance matrix, $\Gamma4, for a model with dependent errors, 
\[
Y=Z \beta+\zeta, \quad \zeta \sim N[0, \Gamma]
\]
$\zeta$ follows a multivariate normal distribution with mean zero and covariance matrix $\Gamma$.

The minimum variance unbiased estimator of $\beta$ for this model is
\[
\hat{\beta}_{G L S}\left(y_{1: N}\right)=\left(Z^{\mathrm{T}} \Gamma^{-1} Z\right)^{-1} Z^{\mathrm{T}} \Gamma^{-1} y
\]
and the variance of $\beta$ is 
\[
\operatorname{Var}\left[\hat{\beta}_{O L S}\left(Y_{1: N}\right)\right]=\left(Z^{\mathrm{T}} Z\right)^{-1} Z^{\mathrm{T}} \Gamma Z\left(Z^{\mathrm{T}} Z\right)^{-1}
\]

The OLS estimator remains unbiased for this model, but it's often a practical solution to use the OLS estimator, expecially for preliminary data analysis.  We  can’t necessarily make a good estimator based on the GLS model *unless* we know $\Gamma$.

It is okay to do ordinary linear regression for data which are not well modeled with uncorrelated errors. However, if we do so, we should not trust the **error estimates** coming from L1.


